\begin{flushleft}
    {\Huge\bfseries Abstract}
\end{flushleft}

\addcontentsline{toc}{chapter}{Abstract}
\vspace{1.5cm}

This work presents a novel approach for mental health and emotion detection from social media
text, addressing the limitations of conventional Natural Language Processing (NLP) models in
handling ambiguity and contextual uncertainty. Traditional models based on TF-IDF and
classification algorithms often fail to capture nuanced emotional states, particularly in
sentences with overlapping or unclear sentiment.

To overcome this, the proposed system integrates Topological Data Analysis (TDA) with machine learning to analyze the geometric structure of text embeddings. By applying persistent homology on dense semantic vectors, the model identifies structural patterns that indicate ambiguity in emotional expression. The classification pipeline is built using sentence embeddings and TF-IDF features combined with an ensemble of XGBoost models, ensuring high classification performance while maintaining interpretability.

A key contribution of this work is the introduction of an ambiguity detection mechanism,
enabling the system to flag uncertain emotional predictions rather than forcing incorrect
classifications. Experimental results demonstrate that while the baseline model achieves
moderate performance, the integration of TDA enhances the system's ability to identify
ambiguous inputs and improves overall robustness.

This approach provides a foundation for more reliable mental health monitoring systems, where
understanding uncertainty is as critical as prediction accuracy.
