\documentclass[10pt]{beamer}

% Modern theme configuration
\usetheme{Madrid}
\usecolortheme{orchid}
\usefonttheme{professionalfonts}

% Custom Colors
\definecolor{DeepNavy}{RGB}{0, 33, 71}
\setbeamercolor{palette primary}{bg=DeepNavy, fg=white}
\setbeamercolor{palette secondary}{bg=DeepNavy!80, fg=white}
\setbeamercolor{palette tertiary}{bg=DeepNavy!60, fg=white}
\setbeamercolor{titlelike}{parent=palette primary}
\setbeamercolor{structure}{fg=DeepNavy}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, calc}
\usepackage{graphicx}

% Metadata
\title[Emotion Detection with TDA]{Emotion Detection from Social Media Posts \\ with Ambiguity Detection}
\subtitle{Using Topological Data Analysis (TDA)}
\author[Arjeet Singh]{Arjeet Singh}
\institute[IIIT-S]{Indian Institute of Information Technology Senapati, Manipur}
\date{April 2026}

\begin{document}

% Title Page
\begin{frame}[plain]
    \titlepage
\end{frame}

% Slide 1
\begin{frame}{Introduction, Objective and Contribution}
    \textbf{Introduction}
    \begin{itemize}
        \item Emotion detection is critical for understanding social media sentiment.
        \item Traditional models force every input into a fixed class, even when ambiguous.
    \end{itemize}

    \vspace{0.3cm}
    \textbf{Objective}
    \begin{itemize}
        \item Develop a system that can explicitly flag \textbf{ambiguous} inputs.
        \item Use Topological Data Analysis (TDA) to quantify semantic uncertainty.
    \end{itemize}

    \vspace{0.3cm}
    \textbf{Contribution}
    \begin{itemize}
        \item Integration of TDA persistence landscapes with BERT and XGBoost.
        \item Novel ambiguity detection rules based on geometric "loop count" (H1).
        \item Robustness against "confident-but-wrong" failure modes in NLP.
    \end{itemize}
\end{frame}

% Slide 2
\begin{frame}{Background / Existing Work}
    \textbf{Existing Work}
    \begin{itemize}
        \item BERT and Sentence-BERT models are state-of-the-art for emotion classification.
        \item Datasets like GoEmotions provide fine-grained labels (28 categories).
    \end{itemize}

    \vspace{0.2cm}
    \textbf{Problems with Existing Work}
    \begin{itemize}
        \item \textbf{Forced Choice:} Models pick a class even for nonsensical or mixed-emotion text.
        \item \textbf{Overconfidence:} Neural networks are often overconfident on unseen samples.
    \end{itemize}

    \vspace{0.2cm}
    \textbf{Why This Project is Needed}
    \begin{itemize}
        \item Real-world data is messy; sarcasm and semantic overlaps create genuine ambiguity.
        \item Identifying uncertainty is critical for reliable AI systems.
    \end{itemize}

    \vspace{0.2cm}
    \textbf{The Solution}
    \begin{itemize}
        \item TDA analyzes the \textbf{geometry} of the training neighborhood.
        \item Complex shapes (loops) in the neighborhood signal high uncertainty.
    \end{itemize}
\end{frame}

% Slide 3
\begin{frame}{Proposed System and Architecture}
    \begin{columns}[T]
        \begin{column}{0.45\textwidth}
            \textbf{System Design}
            \begin{center}
                \resizebox{!}{0.72\textheight}{
                \begin{tikzpicture}[
                  font=\tiny,
                  node distance=0.5cm,
                  box/.style={draw, rectangle, rounded corners=2pt, thick, align=center, text width=2.2cm, fill=blue!5},
                  proc/.style={draw, rectangle, rounded corners=2pt, thick, align=center, text width=2.2cm, fill=green!5},
                  amb/.style={draw, rectangle, rounded corners=2pt, thick, align=center, text width=2.2cm, fill=orange!10},
                  out/.style={draw, rectangle, rounded corners=2pt, thick, align=center, text width=2.2cm, fill=red!5},
                  arr/.style={-{Stealth[length=3pt]}, thick}
                ]
                  \node[box] (in) {\textbf{Input Text}};
                  \node[proc, below=of in] (emb) {\textbf{Sentence-BERT}\\[1pt]Embedding};
                  \node[proc, left=1.0cm of emb, yshift=-0.8cm] (tfidf) {\textbf{TF-IDF}\\[1pt]Features};
                  \node[proc, right=1.0cm of emb, yshift=-0.4cm] (knn) {\textbf{K-Nearest}\\[1pt]Neighbours};
                  \node[proc, below=of knn] (tda) {\textbf{Giotto-TDA}\\[1pt]Persistence};
                  \node[box, below=2.2cm of emb] (fuse) {\textbf{Feature Fusion}};
                  \node[proc, below=of fuse] (xgb) {\textbf{XGBoost}\\[1pt]Classifiers};
                  \node[amb, below=of xgb] (det) {\textbf{Ambiguity}\\[1pt]Detection Rules};
                  \node[out, below=of det] (res) {\textbf{Final Result}\\[1pt](Emotion/Ambiguity)};

                  \draw[arr] (in) -- (emb);
                  \draw[arr] (emb) -| (tfidf);
                  \draw[arr] (emb) -| (knn);
                  \draw[arr] (emb) -- (fuse);
                  \draw[arr] (knn) -- (tda);
                  \draw[arr] (tfidf) |- (fuse);
                  \draw[arr] (tda) |- (fuse);
                  \draw[arr] (fuse) -- (xgb);
                  \draw[arr] (xgb) -- (det);
                  \draw[arr] (det) -- (res);
                \end{tikzpicture}
                }
            \end{center}
        \end{column}
        
        \begin{column}{0.55\textwidth}
            \small
            \textbf{1. Steps (Input to Output)}
            \begin{itemize}
                \item \textbf{Input:} Raw Reddit comment.
                \item \textbf{Process:} Semantic encoding + Geometric analysis.
                \item \textbf{Output:} Labelled emotion or Ambiguity flag.
            \end{itemize}

            \vspace{0.1cm}
            \textbf{2. Tools \& Technology}
            \begin{itemize}
                \item \textbf{Language:} Python 3.10+
                \item \textbf{Libraries:} Giotto-TDA, XGBoost, Transformers (Sentence-BERT).
            \end{itemize}

            \vspace{0.1cm}
            \textbf{3. Working Principle}
            \begin{enumerate}
                \item Retrieve training-set neighbours for the input text.
                \item Analyze neighborhood topology (persistent homology).
                \item Use "Loop Count" and "Confidence Gap" to flag high-uncertainty inputs.
            \end{enumerate}
        \end{column}
    \end{columns}
\end{frame}

% Slide 4
\begin{frame}{Result and Analysis}
    \textbf{Experimental Setup}
    \begin{itemize}
        \item \textbf{Dataset:} GoEmotions (43,410 training samples, 5,427 test).
        \item \textbf{Evaluation:} Accuracy and Macro F1-score for 3-class mapping.
    \end{itemize}

    \vspace{0.2cm}
    \textbf{Quantitative Results}
    \begin{table}
        \centering
        \small
        \begin{tabular}{lccc}
            \toprule
            \textbf{Model Version} & \textbf{Accuracy} & \textbf{Macro F1} & \textbf{Ambiguity Flag} \\
            \midrule
            Embeddings only        & 0.57 & 0.52 & No \\
            Embeddings + TF-IDF    & 0.58 & 0.54 & No \\
            \textbf{Proposed (TDA)} & \textbf{0.59} & \textbf{0.56} & \textbf{Yes} \\
            \bottomrule
        \end{tabular}
        \caption{Performance Comparison Table}
    \end{table}

    \vspace{0.1cm}
    \textbf{Key Finding:}
    TDA features provide independent signals that identify cases where the classifier is uncertain due to neighborhood geometry.
\end{frame}

% Slide 5
\begin{frame}{Conclusion}
    \textbf{Summary of Work}
    \begin{itemize}
        \item Developed a robust emotion detection pipeline for social media text.
        \item Successfully integrated Topological Data Analysis (TDA) to capture geometric uncertainty.
        \item Implemented a multi-rule ambiguity detection layer.
    \end{itemize}

    \vspace{0.4cm}
    \textbf{Future Scope \& Improvements}
    \begin{itemize}
        \item \textbf{Dataset Expansion:} Develop a labeled ambiguity dataset.
        \item \textbf{Optimization:} Explore approximate persistence computation.
        \item \textbf{Multilingual:} Extend to global sentiment analysis.
    \end{itemize}
\end{frame}

\end{document}
